Equivalence Testing and the Second Generation \emph{P}-Value

### 2019-03-08 manuscript.R1.tex
First complete reply to review DL.

\documentclass[floatsintext,man]{apa6}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

% Table formatting
\usepackage{longtable, booktabs}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}

\newenvironment{lltable}
  {\begin{landscape}\begin{center}\begin{ThreePartTable}}
  {\end{ThreePartTable}\end{center}\end{landscape}}

% The following enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand\getlongtablewidth{%
 \begingroup
  \ifcsname LT@\roman{LT@tables}\endcsname
  \global\longtablewidth=0pt
  \renewcommand\LT@entry[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}%
  \@nameuse{LT@\roman{LT@tables}}%
  \fi
\endgroup}

  \usepackage{graphicx}
  \makeatletter
  \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
  \def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
  \makeatother
  % Scale images if necessary, so that they will not overflow the page
  % margins by default, and it is still possible to overwrite the defaults
  % using explicit options in \includegraphics[width, height, ...]{}
  \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            pdfauthor={},
            pdftitle={Equivalence Testing and the Second Generation P-Value},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setlength{\parindent}{0pt}
%\setlength{\parskip}{0pt plus 0pt minus 0pt}

\setlength{\emergencystretch}{3em}  % prevent overfull lines

% Manuscript styling
\captionsetup{font=singlespacing,justification=justified}
\usepackage{csquotes}
\usepackage{upgreek}

\usepackage{tikz} % Variable definition to generate author note

% fix for \tightlist problem in pandoc 1.14
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Essential manuscript parts
  \title{Equivalence Testing and the Second Generation \emph{P}-Value}

  \shorttitle{TOST vs.~SGPV}

  \author{Daniël Lakens\textsuperscript{1}~\& Marie Delacre\textsuperscript{2}}

  \def\affdep{{"", ""}}%
  \def\affcity{{"", ""}}%

  \affiliation{
    \vspace{0.5cm}
          \textsuperscript{1} Eindhoven University of Technology, Eindhoven, The Netherlands\\
          \textsuperscript{2} Service of Analysis of the Data, Université Libre de Bruxelles, Belgium  }

  \authornote{
    \newcounter{author}
    All code associated with this article, including the reproducible
    manuscript, is available from
    \url{https://github.com/Lakens/TOST_vs_SGPV}. This work was supported by
    the Netherlands Organization for Scientific Research (NWO) VIDI grant
    452-17-013.

                      Correspondence concerning this article should be addressed to Daniël Lakens, Den Dolech 1, IPO 1.33, 5600 MB, Eindhoven, The Netherlands. E-mail: \href{mailto:D.Lakens@tue.nl}{\nolinkurl{D.Lakens@tue.nl}}
                          }

  \abstract{To move beyond the limitations of null-hypothesis tests, statistical
approaches have been developed where the observed data is compared
against a range of values that are equivalent to the absence of a
meaningful effect. Specifying a range of values around zero allows
researchers to statistically reject the presence of effects large enough
to matter, and prevents practically insignificant effects from being
interpreted as a statistically significant difference. We compare the
behavior of the recently proposed second generation \emph{p}-value
(Blume, D'Agostino McGowan, Dupont, \& Greevy, 2018) with the more
established Two One-Sided Tests (TOST) equivalence testing procedure
(Schuirmann, 1987). We show that the two approaches yield almost
identical results under optimal conditions. Under suboptimal conditions
(e.g., when the confidence interval is wider than the equivalence range,
or when confidence intervals are asymmetric) the second generation
\emph{p}-value becomes difficult to interpret as a descriptive
statistic. The second generation \emph{p}-value is interpretable in a
dichotomous manner (i.e., when the SGPV equals 0 or 1 because the
confidence intervals lies completely within or outside of the
equivalence range), but this dichotomous interpretation does not require
calculations. We conclude that equivalence tests yield more consistent
\emph{p}-values, distinguish between datasets that yield the same second
generation \emph{p}-value, and allow for easier control of Type I and
Type II error rates.}
  \keywords{equivalence testing, second generation \emph{p}-values, hypothesis
testing, TOST, statistical inference \\

@@

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{example}{Example}
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

\maketitle

\setcounter{secnumdepth}{0}

To test predictions researchers predominantly rely on null-hypothesis
tests.\\
This statistical approach can be used to examine whether observed data
is sufficiently surprising under the null hypothesis to reject an effect
that equals exactly zero. Null-hypothesis tests have an important
limitation, in that this procedure can only reject the hypothesis that
there is no effect, while scientists should also be able to provide
statistical support for \emph{equivalence}. When testing for equivalence
researchers aim to examine whether an observed effect is too small to be
considered meaningful, and therefore is practically equivalent to zero.
By specifying a range around the null hypothesis of values that are
deemed practically equivalent to the absence of an effect (i.e., 0 ±
0.3) the observed data can be compared against an \emph{equivalence
range} and researchers can test if a meaningful effect is absent (Hauck
\& Anderson, 1984; Kruschke, 2018; Rogers, Howard, \& Vessey, 1993;
Serlin \& Lapsley, 1985; Spiegelhalter, Freedman, \& Parmar, 1994;
Wellek, 2010; Westlake, 1972).

Second generation \emph{p}-values (SGPV) were recently proposed to as a
descriptive statistic that represents \enquote{the proportion of
data-supported hypotheses that are also null hypotheses} (Blume et al.,
2018). The researcher specifies an equivalence range around a null
hypothesis of values that are considered practically equivalent to the
null hypothesis. The SGPV measures the degree to which a set of
data-supported parameter values falls within the interval null
hypothesis. If the estimation interval falls completely within the
equivalence range, the SGPV is 1. If the confidence interval falls
completely outside of the equivalence range, the SGPV is 0. Otherwise
the SGPV is a value between 0 and 1 that expresses the overlap of
data-supported hypotheses and the equivalence range. When calculating
the SGPV the set of data-supported parameter values can be represented
by a confidence interval (CI), although one could also choose to use
credible intervals or Likelihood support intervals (SI). When a
confidence interval is used, the SGPV and equivalence tests such as the
Two One-Sided Tests (TOST) procedure (Lakens, 2017; Meyners, 2012;
Quertemont, 2011; Schuirmann, 1987) appear to have close ties, because
both tests compare a confidence interval against an equivalence range.
Here, we aim to examine the similarities and differences between the
TOST procedure and the SGPV.

The TOST procedure also relies on the confidence interval around the
effect. In the TOST procedure the data is tested against the lower
equivalence bound in the first one-sided test, and against the upper
equivalence bound in the second one-sided test (Lakens, Scheel, \&
Isager, 2018). If both tests statistically reject an effect as extreme
or more extreme than the equivalence bound, you can conclude the
observed effect is practically equivalent to zero from a Neyman-Pearson
approach to statistical inferences. Because one-sided tests are
performed, one can also conclude equivalence by checking whether the
1-2\(\times\)\(\alpha\) confidence interval (e.g., when the alpha level
is 0.05, a 90\% CI) falls completely within the equivalence bounds.
Because both equivalence tests as the SGPV are based on whether and how
much a confidence interval overlaps with equivalence bounds, it seems
worthwhile to compare the behavior of the newly proposed SGPV to
equivalence tests to examine the unique contribution of the SGPV to the
statistical toolbox.

@@

The second generation \emph{p}-value (SGPV) is calculated as: \[
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
\] where I is the interval based on the data (e.g., a 95\% confidence
interval) and H\textsubscript{0} is the equivalence range. The first
term of this formula implies that the second generation \emph{p}-value
is the width of the confidence interval that overlaps with the
equivalence range, divided by the total width of the confidence
interval. The second term is a \enquote{small sample correction} (which
will be discussed later) that comes into play whenever the confidence
interval is more than twice as wide as the equivalence range. To examine
the relation between the TOST \emph{p}-value and the SGPV we can
calculate both statistics across a range of observed effect sizes. In
Figure \ref{fig:TOSTSGPV1} \emph{p}-values are plotted for the TOST
procedure and the SGPV. The statistics are calculated for hypothetical
one-sample \emph{t}-tests for observed means ranging from 140 to 150 (on
the x-axis). The equivalence range is set to 145 ± 2 (i.e., an
equivalence range from 143 to 147), the observed standard deviation is
assumed to be 2, and the sample size is 30. For example, for the
left-most point in Figure \ref{fig:TOSTSGPV1} the SGPV and the TOST
\emph{p}-value is calculated for a hypothetical study with a sample size
of 30, an observed standard deviation of 2, and an observed mean of 140,
where the \emph{p}-value for the equivalence test is 1, and the SGPV is
0. Our conclusions about the relationship between TOST \emph{p}-values
and SGPV in this article are not dependent upon any specific example, as
readers can explore for themselves in an online Shiny app:
\url{http://shiny.ieis.tue.nl/TOST_vs_SGPV/}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV1-1.pdf}
\caption{\label{fig:TOSTSGPV1}Comparison of \emph{p}-values from TOST (black
line) and SGPV (dotted grey line) across a range of observed sample
means (x-axis) tested against a mean of 145 in a one-sample
\emph{t}-test with a sample size of 30 and a standard deviation of 2.}
\end{figure}

The SGPV treats the equivalence range as the null-hypothesis, while the
TOST procedure treats the values outside of the equivalence range as the
null-hypothesis. For ease of comparison we can reverse the SGPV (by
calculating 1-SGPV in Figure \ref{fig:TOSTSGPV2}) to make the values
more easily comparable. We see that the \emph{p}-value from the TOST
procedure and the SGPV follow each other closely.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV2-1.pdf}
\caption{\label{fig:TOSTSGPV2}Comparison of \emph{p}-values from TOST (black
line) and 1-SGPV (dotted grey line) across a range of observed sample
means (x-axis) tested against a mean of 145 in a one-sample
\emph{t}-test with a sample size of 30 and a standard deviation of 2.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV3-1.pdf}
\caption{\label{fig:TOSTSGPV3}Means, normal distribution, and 95\% CI for
three example datasets that illustrate the relationship between
\emph{p}-values from TOST and SGPV.}
\end{figure}

Two other points always have to overlap. When the 95\% CI falls
completely inside the equivalence region, and one endpoint of the
confidence interval is exactly equal to one of the equivalence bounds
(see situation B in Figure \ref{fig:TOSTSGPV3}) the TOST \emph{p}-value
(which relies on a one-sided test) is always 0.025. The third point
where the SGPV and the \emph{p}-value from the TOST procedure should
overlap is where the 95\% CI falls completely outside of the equivalence
range, but one endpoint of the confidence interval is equal to the
equivalence bound (see situation C in Figure \ref{fig:TOSTSGPV3}), when
the \emph{p}-value will always be 0.975. Note that this situation is in
essence a minimum-effect test (Murphy, Myors, \& Wolach, 2014). The goal
of a minimum-effect is not just to reject a difference of zero, but to
reject the smallest effect size of interest (i.e., the equivalence
bounds). The SGPV summarizes the information from an equivalence test
and a minimum-effect test. These can be two relevant questions to ask,
although it often makes sense to combine an equivalence test and a
null-hypothesis test instead (Lakens et al., 2018).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV4-1.pdf}
\caption{\label{fig:TOSTSGPV4}Means, normal distribution, and 95\% CI for
samples where the observed population mean is 1.5, 1.4, 1.3, and 1.2.}
\end{figure}

For example, the SGPV from A to D is 0.76, 0.81, 0.86, and 0.91. The
difference in the percentage of overlap between A and B (-0.05) is
identical to the difference in the percentage of overlap between C and D
as the mean gets 0.1 closer to the test value (-0.05). As the observed
mean in a one-sample \emph{t}-test lies closer to the test value (in
Figure \ref{fig:TOSTSGPV4}), from situation A to D, the mean gets closer
to the test value by 0.1) the difference in the overlap changes
uniformly. As we move the observed mean closer to the test value in
steps of 0.1 across A to D the \emph{p}-value calculated for normally
distributed data is not uniformly distributed. The probability of
observing data more extreme than the upper bound of 2 is (from A to D)
0.16, 0.12, 0.08, and 0.06. As we can see, the difference between A and
B (0.04) is not the same as the difference between C and D (0.03).
Indeed, the difference in \emph{p}-values is the largest as you start at
\emph{p} = 0.5 (when the observed mean falls on the test value), which
is why the line in Figure \ref{fig:TOSTSGPV1} is the steepest at
\emph{p} = 0.5. Note that where the SGPV reaches 1 or 0, \emph{p}-values
closely approximate 0 and 1, but never reach these values.

\subsection{When are the SGPV and Equivalence Test
Unrelated?}\label{when-are-the-sgpv-and-equivalence-test-unrelated}

There are three situations where \emph{p}-values from TOST and SGPV are
unrelated. The first two situations were discussed before and can be
seen in Figure \ref{fig:TOSTSGPV1}. When the SGPV is either 0 or 1
\emph{p}-values from the equivalence test fall between 0.975 and 1 or
between 0 and 0.025. Because \emph{p}-values approach 0 or 1, but are
never exactly 0 or 1, and the SGPV is exactly 0 or 1, the two statistics
are completely unrelated within these ranges. The easiest way to see
this is by plotting the SGPV against the \emph{p}-value from the TOST
procedure. The situations where the SPGV and \emph{p}-values from the
TOST procedure are unrelated are indicated by the parts of the curve
where there are vertical straight lines at second generation
\emph{p}-values of 0 and 1.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV5-1.pdf}
\caption{\label{fig:TOSTSGPV5}The relationship between \emph{p}-values from
the TOST procedure and the SGPV for the same scenario as in Figure
\ref{fig:TOSTSGPV1}.}
\end{figure}

>>>>

\documentclass[,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Equivalence Testing and the Second Generation P-Value},
            pdfauthor={Daniël Lakens~\& Marie Delacre},
            pdfkeywords={equivalence testing, second generation \emph{p}-values, hypothesis
testing, TOST, statistical inference},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

  \title{Equivalence Testing and the Second Generation \emph{P}-Value}
    \author{Daniël Lakens\textsuperscript{1}~\& Marie Delacre\textsuperscript{2}}
    \date{}

\shorttitle{TOST vs. SGPV}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Eindhoven University of Technology, Eindhoven, The Netherlands\\\textsuperscript{2} Service of Analysis of the Data, Université Libre de Bruxelles, Belgium}
\keywords{equivalence testing, second generation *p*-values, hypothesis testing, TOST, statistical inference\newline\indent Word count: }
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage[flushleft]{threeparttable}
\usepackage{threeparttablex}

\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

\usepackage{lineno}

\linenumbers

\authornote{All code associated with this article, including
the reproducible manuscript, is available from
\url{https://github.com/Lakens/TOST_vs_SGPV}. This work was supported by
the Netherlands Organization for Scientific Research (NWO) VIDI grant
452-17-013.

Correspondence concerning this article should be addressed to Daniël
Lakens, Den Dolech 1, IPO 1.33, 5600 MB, Eindhoven, The Netherlands.
E-mail:
\href{mailto:D.Lakens@tue.nl}{\nolinkurl{D.Lakens@tue.nl}}}

\abstract{
To move beyond the limitations of null-hypothesis tests, statistical
approaches have been developed where the observed data is compared
against a range of values that are equivalent to the absence of a
meaningful effect. Specifying a range of values around zero allows
researchers to statistically reject the presence of effects large enough
to matter, and prevents practically insignificant effects from being
interpreted as a statistically significant difference. We compare the
behavior of the recently proposed second generation \emph{p}-value
(Blume, D'Agostino McGowan, Dupont, \& Greevy, 2018) with the more
established Two One-Sided Tests (TOST) equivalence testing procedure
(Schuirmann, 1987). We show that the two approaches yield almost
identical results under optimal conditions. Under suboptimal conditions
(e.g., when the confidence interval is wider than the equivalence range,
or when confidence intervals are asymmetric) the second generation
\emph{p}-value becomes difficult to interpret as a descriptive
statistic. The second generation \emph{p}-value is interpretable in a
dichotomous manner (i.e., when the SGPV equals 0 or 1 because the
confidence intervals lies completely within or outside of the
equivalence range), but this dichotomous interpretation does not require
calculations. We conclude that equivalence tests yield more consistent
\emph{p}-values, distinguish between datasets that yield the same second
generation \emph{p}-value, and allow for easier control of Type I and
Type II error rates.

@@

\begin{document}
\maketitle

To test predictions researchers predominantly rely on null-hypothesis
tests. This statistical approach can be used to examine whether observed
data is sufficiently surprising under the null hypothesis to reject an
effect that equals exactly zero. Null-hypothesis tests have an important
limitation, in that this procedure can only reject the hypothesis that
there is no effect, while scientists should also be able to provide
statistical support for \emph{equivalence}. When testing for equivalence
researchers aim to examine whether an observed effect is too small to be
considered meaningful, and therefore is practically equivalent to zero.
By specifying a range around the null hypothesis of values that are
deemed practically equivalent to the absence of an effect (i.e., 0 ±
0.3) the observed data can be compared against an \emph{equivalence
range} and researchers can test if a meaningful effect is absent (Hauck
\& Anderson, 1984; Kruschke, 2018; Rogers, Howard, \& Vessey, 1993;
Serlin \& Lapsley, 1985; Spiegelhalter, Freedman, \& Parmar, 1994;
Wellek, 2010; Westlake, 1972).

Second generation \emph{p}-values (SGPV) were recently proposed to as a
descriptive statistic that represents \enquote{the proportion of
data-supported hypotheses that are also null hypotheses} (Blume et al.,
2018). The researcher specifies an equivalence range around a null
hypothesis of values that are considered practically equivalent to the
null hypothesis. The SGPV measures the degree to which a set of
data-supported parameter values falls within the interval null
hypothesis. If the estimation interval falls completely within the
equivalence range, the SGPV is 1. If the confidence interval falls
completely outside of the equivalence range, the SGPV is 0. Otherwise
the SGPV is a value between 0 and 1 that expresses the overlap of
data-supported hypotheses and the equivalence range. When calculating
the SGPV the set of data-supported parameter values can be represented
by a confidence interval (CI), although one could also choose to use
credible intervals or Likelihood support intervals (SI). When a
confidence interval is used, the SGPV and equivalence tests such as the
Two One-Sided Tests (TOST) procedure (Lakens, 2017; Meyners, 2012;
Quertemont, 2011; Schuirmann, 1987) appear to have close ties, because
both tests compare a confidence interval against an equivalence range.
Here, we aim to examine the similarities and differences between the
TOST procedure and the SGPV. We limit our analysis to normally
distributed continuous data.

The TOST procedure also relies on the confidence interval around the
effect. In the TOST procedure the data is tested against the lower
equivalence bound in the first one-sided test, and against the upper
equivalence bound in the second one-sided test (Lakens, Scheel, \&
Isager, 2018). For an excellent discussion of the strengths and
weaknesses of different frequentist equivalence tests, including
alternatives to the TOST procedure, see (Meyners, 2012). If both tests
statistically reject an effect as extreme or more extreme than the
equivalence bound, you can conclude the observed effect is practically
equivalent to zero from a Neyman-Pearson approach to statistical
inferences. Because one-sided tests are performed, one can also conclude
equivalence by checking whether the 1-2\(\times\)\(\alpha\) confidence
interval (e.g., when the alpha level is 0.05, a 90\% CI) falls
completely within the equivalence bounds. Because both equivalence tests
as the SGPV are based on whether and how much a confidence interval
overlaps with equivalence bounds, it seems worthwhile to compare the
behavior of the newly proposed SGPV to equivalence tests to examine the
unique contribution of the SGPV to the statistical toolbox.

@@

The second generation \emph{p}-value (SGPV) is calculated as: \[
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
\] where I is the interval based on the data (e.g., a 95\% confidence
interval) and H\textsubscript{0} is the equivalence range. The first
term of this formula implies that the second generation \emph{p}-value
is the width of the confidence interval that overlaps with the
equivalence range, divided by the total width of the confidence
interval. The second term is a \enquote{small sample correction} (which
will be discussed later) that comes into play whenever the confidence
interval is more than twice as wide as the equivalence range. To examine
the relation between the TOST \emph{p}-value and the SGPV we can
calculate both statistics across a range of observed effect sizes. In
Figure \ref{fig:TOSTSGPV1} \emph{p}-values are plotted for the TOST
procedure and the SGPV. The statistics are calculated for hypothetical
one-sample \emph{t}-tests for observed means ranging from 140 to 150 (on
the x-axis). The equivalence range is set to 145 ± 2 (i.e., an
equivalence range from 143 to 147), the observed standard deviation is
assumed to be 2, and the sample size is 30. For example, for the
left-most point in Figure \ref{fig:TOSTSGPV1} the SGPV and the TOST
\emph{p}-value is calculated for a hypothetical study with a sample size
of 30, an observed standard deviation of 2, and an observed mean of 140,
where the \emph{p}-value for the equivalence test is 1, and the SGPV is
0. Our conclusions about the relationship between TOST \emph{p}-values
and SGPV hold for an SGPV calculated from confidence intervals, and
assuming continuous normally distributed data. Readers can explore the
relationship between TOST \emph{p}-values and SGPV for themselves in an
online Shiny app: \url{http://shiny.ieis.tue.nl/TOST_vs_SGPV/}.

\begin{figure}
\centering
\includegraphics{manuscript.R1_files/figure-latex/TOSTSGPV1-1.pdf}
\caption{\label{fig:TOSTSGPV1}Comparison of \emph{p}-values from TOST (black
line) and SGPV (dotted grey line) across a range of observed sample
means (x-axis) tested against a mean of 145 in a one-sample
\emph{t}-test with a sample size of 30 and a standard deviation of 2.}
\end{figure}

The SGPV treats the equivalence range as the null-hypothesis, while the
TOST procedure treats the values outside of the equivalence range as the
null-hypothesis. For ease of comparison we can plot 1-SGPV (see Figure
\ref{fig:TOSTSGPV2}) to make the values more easily comparable. We see
that the \emph{p}-value from the TOST procedure and the SGPV follow each
other closely. When we discuss the relationship between the p-values
from TOST and the SGPV, we focus on their correspondence at three
values, namely where the TOST \emph{p} = 0.025 and SGPV is 1, where the
TOST \emph{p} = 0.5 and SGPV = 0.5, and where the TOST \emph{p} = 0.975
and SGPV = 1. These three values are important for the SGPV because they
indicate the values at which the SGPV indicates the data should be
interpreted as compatible with the null hypothesis (SGPV = 1), or with
the alternative hypothesis (SGPV = 0), or when the data are strictly
inconclusive (SGPV = 0.5). These three points of overlap are indicated
by the horizontal dotted lines in Figure \ref{fig:TOSTSGPV2}.

\begin{figure}
\centering
\includegraphics{manuscript.R1_files/figure-latex/TOSTSGPV2-1.pdf}
\caption{\label{fig:TOSTSGPV2}Comparison of \emph{p}-values from TOST (black
line) and 1-SGPV (dotted grey line) across a range of observed sample
means (x-axis) tested against a mean of 145 in a one-sample
\emph{t}-test with a sample size of 30 and a standard deviation of 2.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript.R1_files/figure-latex/TOSTSGPV3-1.pdf}
\caption{\label{fig:TOSTSGPV3}Means, normal distribution, and 95\% CI for
three example datasets that illustrate the relationship between
\emph{p}-values from TOST and SGPV.}
\end{figure}

Two other points always have to overlap. When the 95\% CI falls
completely inside the equivalence region, and one endpoint of the
confidence interval is exactly equal to one of the equivalence bounds
(see situation B in Figure \ref{fig:TOSTSGPV3}) the TOST \emph{p}-value
(which relies on a one-sided test) is always 0.025, and the SGPV is 1.
The third point where the SGPV and the \emph{p}-value from the TOST
procedure should overlap is where the 95\% CI falls completely outside
of the equivalence range, but one endpoint of the confidence interval is
equal to the equivalence bound (see situation C in Figure
\ref{fig:TOSTSGPV3}), when the \emph{p}-value will always be 0.975, and
the SGPV is 0. Note that this situation is in essence a minimum-effect
test (Murphy, Myors, \& Wolach, 2014). The goal of a minimum-effect is
not just to reject a difference of zero, but to reject the smallest
effect size of interest (i.e., the equivalence bounds). An equivalence
test and minimum effect test against the same equivalence bound are
complementary, and when a TOST \emph{p}-value is larger than 0.975, the
\emph{p}-value for the minimum effect test is smaller than 0.05, and the
minimum effect test provides no additional information that can not be
derived from the \emph{p}-value from the equivalence test. The SGPV
summarizes the information from an equivalence test and a minimum-effect
test. These can be two relevant questions to ask, although it often
makes sense to combine an equivalence test and a null-hypothesis test
instead (Lakens et al., 2018).

\begin{figure}
\centering
\includegraphics{manuscript.R1_files/figure-latex/TOSTSGPV4-1.pdf}
\caption{\label{fig:TOSTSGPV4}Means, normal distribution, and 95\% CI for
samples where the observed population mean is 1.5, 1.4, 1.3, and 1.2.}
\end{figure}

For example, in Figure \ref{fig:TOSTSGPV4} we have plotted four SGPV's.
From A to D the SGPV is 0.76, 0.81, 0.86, and 0.91. The difference in
the percentage of overlap between A and B (-0.05) is identical to the
difference in the percentage of overlap between C and D as the mean gets
0.1 closer to the test value (-0.05). As the observed mean in a
one-sample \emph{t}-test lies closer to the test value, from situation A
to D, the mean gets closer to the test value by 0.1) the difference in
the overlap changes uniformly. As we move the observed mean closer to
the test value in steps of 0.1 across A to D the \emph{p}-value
calculated for normally distributed data is not uniformly distributed.
The probability of observing data more extreme than the upper bound of 2
is (from A to D) 0.16, 0.12, 0.08, and 0.06. As we can see, the
difference between A and B (0.04) is not the same as the difference
between C and D (0.03). Indeed, the difference in \emph{p}-values is the
largest as you start at \emph{p} = 0.5 (when the observed mean falls on
the test value), which is why the line in Figure \ref{fig:TOSTSGPV1} is
the steepest at \emph{p} = 0.5. Note that where the SGPV reaches 1 or 0,
\emph{p}-values closely approximate 0 and 1, but never reach these
values.

\subsection{\texorpdfstring{When different \emph{p}-values for
equivalence tests yield the same
SGPV}{When different p-values for equivalence tests yield the same SGPV}}\label{when-different-p-values-for-equivalence-tests-yield-the-same-sgpv}

There are three situations where \emph{p}-values for TOST differentiate
between observed results, while the SGPV does not differentiate. The
first two situations were discussed before and can be seen in Figure
\ref{fig:TOSTSGPV1}. When the SGPV is either 0 or 1, \emph{p}-values
from the equivalence test fall between 0.975 and 1 or between 0 and
0.025. While the SGPV is 1 as long as the confidence interval falls
completely within the equivalence bounds, the \emph{p}-value for the
TOST continues to differentiate between results as a function of how far
the confidence interval lies within the equivalence bounds (the further
the confidence interval is from both bounds, the lower the
\emph{p}-value). The easiest way to see this is by plotting the SGPV
against the \emph{p}-value from the TOST procedure. The situations where
the \emph{p}-values from the TOST procedure continue to differentiate
based on how extreme the results are, but the SGPV is a fixed value are
indicated by the parts of the curve where there are vertical straight
lines at second generation \emph{p}-values of 0 and 1.

\begin{figure}
\centering
\includegraphics{manuscript.R1_files/figure-latex/TOSTSGPV5-1.pdf}
\caption{\label{fig:TOSTSGPV5}The relationship between \emph{p}-values from
the TOST procedure and the SGPV for the same scenario as in Figure
\ref{fig:TOSTSGPV1}.}
\end{figure}

A third situation in which the SGPV remains stable across a range of
observed effects, while the TOST \emph{p}-value continues to
differentiate, is whenever the CI is wider than the equivalence range,
and the CI overlaps with the upper \emph{and} lower equivalence bound.
When the confidence interval is more than twice as wide as the
equivalence range the SGPV is set to 0.5. Blume et al. (2018) call this
the \enquote{small sample correction factor}. However, it is not a
correction in the typical sense of the word, since the SGPV is not
adjusted to any \enquote{correct} value. When the normal calculation
would be \enquote{misleading} (i.e., the SGPV would be small, which
normally would suggest support for the alternative hypothesis, but at
the same time all values in the equivalence range are supported), the
SGPV is set to 0.5 which according to Blume and colleagues signals that
the SGPV is \enquote{uninformative}. Note that the CI can be twice as
wide as the equivalence range whenever the sample size is small (and the
confidence interval width is large) \emph{or} when then equivalence
range is narrow. It is therefore not so much a \enquote{small sample
correction} as it is an exception to the typical calculation of the SGPV
whenever the ratio of the confidence interval width to the equivalence
range exceeds 2:1 and the CI overlaps with the upper and lower bounds.
